---
title: "E - LULU processing"
author: "Tobias G Frøslev"
date: "11/5/2017" #Updated 28/6/2017
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file records the processing of OTU tables with the LULU algorithm for the manuscript "Reliable biodiversity metrics from co-occurence based post-clustering curation of amplicon data". 20 (unprocessed) OTU tables and their corresponding files with representative sequences (centroids) have been constructed with different algorithms (VSEARCH, SWARM, DADA2 and DADA2+VSEARCH) and are now ready to be curated by the LULU algorithm.    

This step should be carried out after the taxonomic filtering of primary OTU tables documentet in the file: D_Taxonomic_filtering.Rmd    
NB: All markdown chuncks are set to "eval=FALSE". Change these accordingly. Also code blocks to be run outside R, has been #'ed out. Change this accordingly.  

## Running the curation with LULU for all initial OTU tables produced with VSEARCH, SWARM, DADA2 and CROP 
### Bioinformatic tools necessary
Make sure that you have the following bioinformatic tools in your PATH  
VSEARCH v.2.02 or later (https://github.com/torognes/vsearch)
Blastn 2.4.0+ or later blastn v2.4.0+ (ftp://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/LATEST/).  
LULU - (https://github.com/tobiasgf/lulu)

###Provided scripts  
A number of scripts are provided with this manuscript. Place these in you /bin directory and make them executable with "chmod 755 SCRIPTNAME"" or place the scripts in the directory/directories where they should be executed (i.e. the analyses directory)  
Alfa_matchlist.sh  
  
###Analysis files  
This step is dependent on the presence of otutables from the inital rounds.  

Setting directories and libraries etc
```{r, eval=FALSE}
setwd("~/analyses")
main_path <- "~/analyses" 
path <- file.path(main_path, "otutables_processing")
library(lulu)
```

###Producing match lists
Match lists can be produced with different algorithms mathing all centroids against each other. Here we use blastn as this gives a better matching of erroneous sequences caused by "deletions/insertions" and/or sequences of unequal length. This task will take some time, as several of the datasets contain a high number of centroids.
Produce match lists for all centroid files (xxx.plantcentroids)
```{r, eval=FALSE}
# bash Alfa_matchlist.sh
```
Now we have three matching files for each dataset (otutable (XXX.planttable),centroids (XXX.plantcentroids) and a match list (XXX.centroids.matchlist)). Now we can run the LULU algorithm for each otutable from each analyses using the file pairs XXX.planttable and XXX.centroids.matchlist

###Running the LULU algorithm on the datasets
Now we are ready to run the LULU algorithm on the datasets. The commands below uses the LULU algorithm as a function defined in the separate r source file, LULU.r.   
Process the datasets with LULU
```{r, eval=FALSE}
allFiles <- list.files(path)
allTabs <- allFiles[grepl("planttable$", allFiles)]
allMls <- allFiles[grepl("matchlist$", allFiles)]
tab_names <- sort(as.vector(
 sapply(allTabs, function(x) strsplit(x, ".planttable")[[1]][1])))
ml_names <- sort(as.vector(
 sapply(allMls,function(x) strsplit(x,".plantcentroids.matchlist")[[1]][1])))

if (!all(tab_names == ml_names)) {
  stop("not mathing set of otutables and matchlists", call.=FALSE)
}

read_tabs <- file.path(path, allTabs)
read_mls <- file.path(path, allMls)
proc_files <- file.path(path, paste0("LULU-",tab_names))
proc_tabs <- file.path(path, paste0(allTabs,"_luluprocessed"))
# Vector for filtering... at this step redundant, but included for safety
samples <- c("S001","S002","S003","S004","S005","S006","S007","S008","S067",
             "S009","S010","S011","S012","S013","S014","S040","S068","S015",
             "S016","S017","S018","S069","S070","S019","S020","S021","S022",
             "S024","S025","S026","S027","S041","S028","S029","S030","S032",
             "S033","S034","S035","S042","S036","S037","S038","S039","S086",
             "S087","S088","S089","S044","S071","S045","S046","S047","S048",
             "S049","S050","S051","S052","S053","S055","S056","S057","S058",
             "S090","S059","S060","S061","S062","S063","S064","S065","S066",
             "S072","S073","S074","S075","S076","S077","S078","S091","S079",
             "S080","S081","S082","S083","S084","S085","S092","S094","S095",
             "S096","S097","S098","S099","S100","S101","S102","S103","S104",
             "S106","S107","S108","S109","S133","S110","S111","S112","S113",
             "S114","S115","S116","S117","S118","S119","S120","S121","S122",
             "S123","S124","S134","S125","S126","S127","S129","S130","S131",
             "S132","S135","S136","S137")

tab <- list()
ml <- list()

for(i in seq(1:length(read_tabs))) {
  tab[[i]] <- read.csv(read_tabs[i],sep='\t',header=T,as.is=TRUE,row.names = 1)
  tab[[i]] <- tab[[i]][which(rowSums(tab[[i]]) > 0),samples]
  ml[[i]] <- read.csv(read_mls[i],sep='\t',header=F,as.is=TRUE)
  proc_min <- lulu(tab[[i]],ml[[i]]) ## RUNNING LULU for each table
  curated_table <- proc_min$curated_table ## extracting the curated table (line updated 17/01/2017)
  saveRDS(proc_min,proc_files[i])
  {write.table(curated_table, proc_tabs[i], sep="\t",quote=FALSE, 
               col.names = NA)}
}
```
Now we have a processed table ("XXX.planttable_luluprocessed") for each original table ("XXX.planttable"), and these can be compared.
