---
title: "G - Producing Figure 1"
author: "Tobias G Frøslev"
date: "11/5/2017" # Updated 28/6/2017
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file records the graphical visualisation of the table metrics for the manuscript "Reliable biodiversity metrics from co-occurence based post-clustering curation of amplicon data".  

This step should be carried out after the calculation of table metrics/statistics documented in the file: F_Calculating_statistics_for_tables.Rmd   

NB: All markdown chuncks are set to "eval=FALSE". Change these accordingly. Also code blocks to be run outside R, has been #'ed out. Change this accordingly. 

Setting directories and libraries etc
```{r, eval=FALSE}
setwd("~/analyses")
setwd("~/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses")
main_path <- getwd() 
path <- file.path(main_path, "otutables_processing")
library(stringr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpmisc)
library(cowplot)
```

Make display dotplots of plant richness vs OTU richness
```{r,eval=FALSE}
tab_name <- file.path(main_path,"Table_richness_calculations_long.txt")
total_richness_df2 <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)
total_richness_df2$Level <- 
 factor(total_richness_df2$Level,levels = c("99/100", "98.5", "98", "97", 
                                            "96", "95"))
total_richness_df2$Method <- 
 factor(total_richness_df2$Method,levels = c("VSEARCH", "SWARM", "DADA2(+VS)", 
                                             "CROP"))
total_richness_df2$Curated <- 
 factor(total_richness_df2$Curated,levels = c("raw", "curated"))

formula <- y ~ x

#Extract 97% level
level97 <- total_richness_df2 %>% filter(Level == "97")
Plot97_xy <- ggplot(level97, aes(x= Obs, y= OTU, color = Curated)) +
  geom_point(pch=21,size=1, alpha = 0.8) +
  geom_abline(intercept = 0, linetype =2) +
  facet_grid(. ~Method) +
  geom_smooth(method = "lm", se = F,size=0.5) +
  stat_poly_eq(geom = "label", alpha = 0.5,
               aes(label = paste(..eq.label.., ..rr.label..,sep = "~~~")),
               formula = formula, label.x.npc = "left", label.y.npc = 0.9,
               parse = TRUE, size = 2, label.size = NA) +
  scale_color_brewer(palette = "Set1") +
  xlab("Plant richness") +
  ylab("OTU richness") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw()+
  theme(text = element_text(size=7)) + 
  guides(colour = guide_legend(title="")) + 
  theme(legend.position = c(0.83, 0.6))
```

Make plots of method related statistics
```{r,eval=FALSE}
tab_name <- file.path(main_path,"Table_method_statistics.txt")
method_statistics <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)
method_statistics$Level <- 
 factor(method_statistics$Level,levels = c("99/100", "98.5", "98", "97", 
                                           "96", "95"))
method_statistics$Method <- 
 factor(method_statistics$Method,levels = c("VSEARCH", "SWARM", "DADA2(+VS)", 
                                            "CROP"))
method_statistics$Curated <- 
 factor(method_statistics$Curated,levels = c("raw", "curated"))

#Get beta diversity of plant survey
tab_name <- file.path(main_path,"Table_plants_2014_cleaned.txt")
Plant_data2014 <- read.table(tab_name, sep="\t", row.names = 1, header=TRUE, 
                             as.is=TRUE)
Plant_richness <- colSums(Plant_data2014)
obs_beta <- nrow(Plant_data2014)/mean(Plant_richness)

#Extract 97% Level methods
Level97m <- method_statistics %>% filter(Level == "97")

#Plot OTU count at 97%
Plot97_otucount <- ggplot(Level97m,aes(x=Curated,weights=OTU_count,fill=Curated)) +
  geom_bar(position="dodge",width=.5) + 
  geom_hline(yintercept = 564,linetype = 2) +
  facet_grid(.~Method) +
  xlab("") +
  ylab("OTUs") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw()+
  theme(strip.background = element_blank(), strip.text.x = element_blank())+ 
  theme(text = element_text(size=7))

#Plot taxonomic redundancy at 97%
Plot97_taxred <-  ggplot(Level97m,aes(x=Curated,weights=Redundancy,fill=Curated)) +
  geom_bar(position="dodge",width=.5) + 
  facet_grid(.~Method) +
  xlab("") +
  ylab("Redundancy") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw()+
  theme(strip.background = element_blank(), strip.text.x = element_blank())+ 
  scale_y_continuous(labels = scales::percent)  + 
  theme(text = element_text(size=7))

#Plot beta diversity at 97%
Plot97_beta <- ggplot(Level97m,aes(x=Curated,weights=Beta,fill=Curated)) +
  geom_bar(position="dodge",width=.5) + 
  geom_hline(yintercept = obs_beta,linetype = 2) +
  facet_grid(.~Method) +
  xlab("") +
  ylab("Betadiversity") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw()+
  theme(strip.background = element_blank(), strip.text.x = element_blank()) + 
  theme(text = element_text(size=7))

```

Process the match rates (pident, best match on genbank) for curated vs. discarded OTUs for processed tables, and make display violin plots
```{r,eval=FALSE}
tab_name <- file.path(main_path,"Table_OTU_match_rates.txt")
pident_frame <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)

pident_frame$level_pident <- 
 factor(pident_frame$level_pident,levels = c("99/100", "98.5", "98", "97", 
                                             "96", "95"))
pident_frame$method_pident <- 
 factor(pident_frame$method_pident,levels = c("VSEARCH", "SWARM", 
                                              "DADA2(+VS)", "CROP"))

#Get 97% level for all methods
Level97p <- pident_frame %>% filter(level_pident == "97")

#Violin plot of distribution of best matches, 97% level
Plot97_pident_violin <- 
 ggplot(Level97p, aes(x=retained_or_discarded, 
                      y=pident/100,fill=retained_or_discarded)) +
  geom_violin() +
  facet_grid(.~method_pident) +
  xlab("") +
  ylab("Reference match") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
  theme(text = element_text(size=7)) + 
  guides(fill=guide_legend(title="OTUs"))
```

Make display taxonomic dissimilarity plots
```{r, eval=FALSE}
tab_name <- file.path(main_path,"Table_taxonomic_dissimilatiry_long.txt")
gathered_dissimilarity_long <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)

gathered_dissimilarity2 <- 
 spread(gathered_dissimilarity_long,Curated,Dissimilarity)
gathered_dissimilarity2$Level <- 
 factor(gathered_dissimilarity2$Level,levels = c("99/100", "98.5", "98", "97", 
                                                 "96", "95"))
gathered_dissimilarity2$Method <- 
 factor(gathered_dissimilarity2$Method,levels = c("VSEARCH", "SWARM", 
                                                  "DADA2(+VS)", "CROP"))


Level97d <- gathered_dissimilarity2 %>% filter(Level == "97")
Plot97_dissim <- ggplot(Level97d,aes(x=raw,y=curated)) +
  geom_point(size=1,alpha=0.8,pch=21) + 
  facet_grid(. ~Method) +
  geom_abline(intercept = 0, linetype =2) + 
  xlab("Taxonomic dissimilarity before curation") +
  ylab("after") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw() +
  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
  theme(text = element_text(size=7))
```

Make composite plot (Figure 1) for 97% level
```{r,eval=FALSE}
p1 <- Plot97_xy
p2 <- Plot97_otucount + theme(legend.position="none")
p3 <- Plot97_taxred + theme(legend.position="none")
p4 <- Plot97_beta + theme(legend.position="none")
p5 <- Plot97_pident_violin + theme(legend.position="none")
p6 <- Plot97_dissim + theme(legend.position="none")

Composite_plot <- plot_grid(p1,p2,p3,p4,p5,p6, align = "v", 
                            nrow = 6, ncol=1, rel_heights = c(5.5, 2, 2, 2, 3,2),
                            labels = c("a","b","c","d","e","f"))

ggsave("Fig1_rev1.pdf", Composite_plot, width=6, height=11)

```

