---
title: "I - GenBank coverage"
author: "Tobias G Frøslev"
date: "11/6/2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r set-options, echo=FALSE, cache=FALSE}
options(width=80)
```
This file records the test of the number of ingroup taxa present with ITS2 data on GenBank for the manuscript "Reliable biodiversity metrics from co-occurence based post-clustering curation of amplicon data".   

This part can be run at any point.  
NB: All markdown chuncks are set to "eval=FALSE". Change these accordingly. Also code blocks to be run outside R, has been #'ed out. Change this accordingly.  

## Taxonomic filtering (Plants) of primary OTU tables 
### Bioinformatic tools necessary
Make sure that you have the following bioinformatic tools in your PATH  
rentrez r-package (https://cran.r-project.org/web/packages/rentrez/index.html)  

###Provided scripts  
All codes needed for this step are included below  
  
###Analysis files  
A species level species-site matrix of plant survey data (Table_plants_2014_cleaned.txt)  

```{r, eval=FALSE}
setwd("~/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses")
main_path <- getwd()
path <- file.path(main_path, "otutables_processing")
library(rentrez)

tab_name <- file.path("/Users/Tobias/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses/Table_plants_2014_cleaned.txt")
Plant_data2014 <- read.table(tab_name, sep="\t", row.names = 1, header=TRUE,as.is=TRUE)
plant_names <- row.names(Plant_data2014) # 564 in total

results <- list()
hit_numbers <- vector()

for (tax_no in seq_along(plant_names)){
  search_term <- paste0(plant_names[tax_no],"[Organism] AND internal_transcribed_spacer_2[misc_feature]")
  results[[tax_no]] <- entrez_search(db="nuccore", term=search_term, retmax=10)
  hit_numbers[tax_no] <- results[[tax_no]]$count
}

saveRDS(results,"GenBank_taxon_retrievalRDS")
saveRDS(hit_numbers,"GenBank_taxon_hitsRDS")
saveRDS(hit_numbers,"GenBank_taxon_hitsRDS")

hit_numbers <- readRDS("GenBank_taxon_hitsRDS")

#get number of study species with ITS2 data in GenBank 
sum(hit_numbers == 0) # 63
sum(hit_numbers > 0) # 501
501/564 # 88.8%
```
