---
title: "J - DBOTU3 one-step clustering"
author: "Tobias G Frøslev"
date: "11/6/2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r set-options, echo=FALSE, cache=FALSE}
options(width=80)
```
This file records the clustering of sequence data for the manuscript "Reliable biodiversity metrics from co-occurence based post-clustering curation of amplicon data" using dbotu3. First a 0% (100%) OTU table is produced with VSEARCH.    

This part can be run after the initial merging and demultiplexing of samples, documented in: A_Preparation_of_sequences.Rmd  
NB: All markdown chuncks are set to "eval=FALSE". Change these accordingly. Also code blocks to be run outside R, has been #'ed out. Change this accordingly.  

## VSEARCH clustering of plant data
### Bioinformatic tools necessary
Make sure that you have the following bioinformatic tools in your PATH  
VSEARCH v.2.32 or later (https://github.com/torognes/vsearch)  
uc2otutab.py  (see http://drive5.com/python/summary.html)  
dbotu3 (https://github.com/swo/dbotu3)  
BlastN - blastn v2.4.0+ (ftp://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/LATEST/)  

###Provided scripts  
A number of scripts are provided with this manuscript. Place these in you /bin directory and make them executable with "chmod 755 SCRIPTNAME"" or place the scripts in the directory/directories where they should be executed (i.e. the analyses directory)  
Alfa_vsearch_dbotu.sh    
rename.pl  
  
###Analysis files  
A number of files provided with this manuscript are necessary for the processing (they need to be placed in the analyses directory): This script parses the dereplicated sample wise fastafiles produced by the previous step (see Preparation_of_sequences.Rmd)  


Make a 0% clustering table for use with bdotu3
```{r}
# mkdir -p dbotu3
# cp S*.fas dbotu3 
# cd dbotu3
# bash Alfa_vsearch_dbotu.sh
```
Now we have a table (VSEARCH_100.otutable) and a corresponding centroids file (VSEARCH_100.centroids).  

Process the OTU table with dbotu3 using abundance criteion 0 (a=0)
```{r}
# dt dbotuVS100
# python dbotu3.py --dist 0.16 --abund 0 --log VSEARCH_100.log --output VSEARCH_100.otutable_dbotuprocessed0 VSEARCH_100.otutable VSEARCH_100.centroids
```

Process the OTU table with dbotu3 using abundance criteion 10 (a=10)
```{r}
#dt dbotuVS100_2
#python dbotu3.py --dist 0.16 --abund 10 --log VSEARCH_100_a10.log --output VSEARCH_100.otutable_dbotuprocesseda10 VSEARCH_100.otutable VSEARCH_100.centroids```
```

make the files ready for benchmarking against LULU
```{r}
# mkdir -p dbotu_processing
# cp VSEARCH_100.otutable_dbotuprocessed0 dbotu_processing/
# cp VSEARCH_100.otutable_dbotuprocessed10 dbotu_processing/
# cp VSEARCH_100.centroids dbotu_processing/
# cd dbotu_processing
```


Setting directories and libraries etc
```{r, eval=FALSE}
setwd("~/analyses")
setwd("~/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses")
main_path <- getwd()
path <- file.path(main_path, "dbotu_processing")
library(stringr)
library(dplyr)
library(tidyr)
require(vegan)
library(ggplot2)
library(ggpmisc)
library("taxize")
```

Moving files to laptop
```{r}
# scp p:data/Biowide/METHOD_PAPER/analyses_rerun/dbotu3/dbotu_processing/*dbotuprocessed* /Users/Tobias/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses/dbotu_processing
# scp p:data/Biowide/METHOD_PAPER/analyses_rerun/dbotu3/dbotu_processing/VSEARCH_100.centroids /Users/Tobias/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses/dbotu_processing
```

Getting the centroids
```{r}
path <- file.path(main_path, "dbotu_processing")

read_centr <- file.path(path, "VSEARCH_100.centroids")
allcentroids <- read.csv(read_centr,sep='\t',header=F,as.is=TRUE)
otusID <- seq(1,length(allcentroids$V1),2)
seqsID <- seq(2,length(allcentroids$V1),2)
otus <- allcentroids[otusID,]
seqs <- allcentroids[seqsID,]
otus <- gsub(">","",otus)
centroid_df <- data.frame(qseqid = otus, sequence = seqs)

centroid_tab <- file.path(path,"centroids_table_dbotu.txt")
{write.table(centroid_df, centroid_tab, sep="\t",quote=FALSE, col.names = NA)}
```

Extracting the curated OTU sequences:
```{r}
centroid_tab <- file.path(path,"centroids_table_dbotu.txt")
centroid_df <- read.table(centroid_tab, sep="\t", header=TRUE, as.is=TRUE)

#Extract for the a=0 run
tab_name <- file.path(path,"dbotu3_0.otutable")
dbotutable_a0 <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)
tab_name <- file.path(path,"dbotu3_10.otutable")
dbotutable_a10 <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)

all_centroids <- union(dbotutable_a0$OTU_ID,dbotutable_a10$OTU_ID)

ingroup_seqs <- centroid_df[which(centroid_df$qseqid %in% all_centroids),]

sinkname <- file.path(path, "dbotu3_all_centroids.txt")
sink(sinkname)
for (i in seq(1:dim(ingroup_seqs)[1])){
  {header <- paste0(">",ingroup_seqs$qseqid[i],"\n")
   cat(header)
   seqq <- paste0(ingroup_seqs$sequence[i],"\n")
   cat(seqq)
  }
}
sink()

```

Get Blasthits
```{r}
#For a=0 run
# ~/bin/blastn -db nt -num_threads 50 -max_target_seqs 20 -outfmt '6 std qlen ssciname staxid' -out dbotu3_all.blasthits -qcov_hsp_perc 90 -perc_identity 80 -query dbotu3_all_centroids.txt
```

Moving blasthits files to laptop
```{r}
# scp p:data/Biowide/METHOD_PAPER/analyses_rerun/dbotu3/dbotu_processing/dbotu3_all.blasthits /Users/Tobias/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses/dbotu_processing
# scp p:data/Biowide/METHOD_PAPER/analyses_rerun/dbotu3/dbotu_processing/VSEARCH_100.centroids /Users/Tobias/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses/dbotu_processing
```

Reading blasthits file
```{r,eval=FALSE}
IDtable_name <- file.path(path,"dbotu3_all.blasthits")
IDtable=read.csv(IDtable_name,sep='\t',header=F,as.is=TRUE)
names(IDtable) <- c("qseqid","sseqid","pident","length","mismatch","gapopen","qstart","qend","sstart","send","evalue","bitscore","qlen","ssciname","staxid")
```

Filter list of hits so it only contains the top hits for each OTU (top hits defined as the best hit and ~0.50% down, i.e from 100% down to more than 99.49%, or from 97.5% down to more than 96.9%, set by the variable "margin")
```{r,eval=FALSE}
margin <- 0.51
new_IDtable <- IDtable[0,] # prepare filtered matchlist
ids <- names(table(IDtable$qseqid))
i=1
o=length(ids)
for (name in ids){
  print(paste0("progress: ", round(((i/o) * 100),0) ,"%")) # make a progressline
  test <- IDtable[which(IDtable$qseqid == name),] # select all lines for a query
  max <- max(test$pident)
  test <- test[which(test$pident > (max-margin)),] # select all lines for a query
  #These lines can be included if analysing a taxonomic group with a lot of
     #"unassigned" sequences in GenBank, to exclude those from further evaluation.
  #test2 <- test[!grepl("uncultured eukaryote",
  #          test$truncated_ssciname,ignore.case = TRUE),] 
  #if (nrow(test2) > 1) {test <- test2}
  #test <- test[!grepl("Environmental",
  #          test$truncated_ssciname,ignore.case = TRUE),]
  if (nrow(test) > 0 ) { test$string <- toString(names(table(test$ssciname))) }
  new_IDtable = rbind(new_IDtable,test) # add this row to the filtered IDtable
  i=i+1
}
```
Now we have a filtered list with only top hits for each OTU/centroid. We need to calculate the most commonly applied taxonomic annotation for each.  

Calculate the most commonly used taxonomic annotation (taxid) for each OTU
```{r,eval=FALSE}
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Apply function to blasthits
new_IDtable$majority_taxid <-  with(new_IDtable, ave(staxid, qseqid , FUN=Mode))
IDtable2 = new_IDtable[!duplicated(new_IDtable[c(1,17)]),]
```
Now our list only contain the most common taxid for each OTU. We need to get the full taxonomic affiliation to be able to sort at higher taxonomic levels.  

###Get full taxonomic path for each OTU
Get the taxonomic string (kingdom, phylum, class, order, family, genus, species) for each OTU (e.g. k__Viridiplantae;p__Streptophyta;c__Liliopsida;o__Poales;f__Poaceae;g__Agrostis;s__Agrostis_vinealis). Using the r-package taxize.
```{r,eval=FALSE}
all_staxids <- names(table(IDtable2$staxid)) # get all taxids for table
all_classifications <- list() # prepare list for taxize output
o=length(all_staxids) # number of taxids

Start_from <- 1 # change if loop needs to be restarted due to time-out

#Get ncbi classification of each entry
for (cl in Start_from:o){ # the taxize command "classification" can be run on 
  #the all_staxids vector in one line, but often there is
  #a timeout command, therefor this loop workaround.
 
  #make a progressline (indicating the index the loops needs to be
  #restarted from if it quits)
  print(paste0("processing: ", cl , " of ", o , " taxids")) 
  all_classifications[cl] <- classification(all_staxids[cl], db = "ncbi")
}

#Construct a taxonomic path from each classification
output <- data.frame(staxid=character(),taxpath=character(),
                     stringsAsFactors=FALSE)
totalnames <- length(all_staxids)
for (curpart in seq(1:totalnames)){
  print(paste0("progress: ", round(((curpart/totalnames) 
                                    * 100),0) ,"%")) # make a progressline
  currenttaxon <- all_classifications[curpart][[1]]
  if ( !is.na(currenttaxon)) {
    spec <- all_staxids[curpart]
    gen <- currenttaxon[which(currenttaxon$rank == "genus"),"name"]
    fam <- currenttaxon[which(currenttaxon$rank == "family"),"name"]
    ord <- currenttaxon[which(currenttaxon$rank == "order"),"name"]
    cla <- currenttaxon[which(currenttaxon$rank == "class"),"name"]
    phy <- currenttaxon[which(currenttaxon$rank == "phylum"),"name"]
    kin <- currenttaxon[which(currenttaxon$rank == "kingdom"),"name"]
    spe <- currenttaxon[which(currenttaxon$rank == "species"),"name"]
    currentpath <- gsub(" ", "_", 
                        paste0("k__",kin,";p__",phy,";c__",cla,";o__",ord,";f__",fam,";g__",gen,";s__",spe))
    output[curpart,"staxid"] <-  spec # add row to the filtered IDtable
    output[curpart,"taxpath"] <-  currentpath # add row to the filtered IDtable
  }
}
```
...this will give some warnings, which is OK  

Merge the taxonomic string with the filtered hit list, and save the list
```{r,eval=FALSE}
taxonomic_info <- merge(IDtable2,output,by = "staxid", all=TRUE)
tbname <- file.path(path,"Table_otu_taxonomy_dbotu.txt")
{write.table(taxonomic_info, tbname, sep="\t",quote=FALSE, col.names = NA)}
```
Now we have a table ("Table_otu_taxonomy.txt") with "full" taxonomic information for each OTU, that allows us to filter our OTU tables for ingroup.  

###Identify ingroup OTUs
Identify ingroup OTUs (plants) to match the reference dataset (keeping phylum Spermatophyta, but excluding a couple classes of "Bryophytes" and "Algae").
```{r,eval=FALSE}
tbname <- file.path(path,"Table_otu_taxonomy_dbotu.txt")
# reads the table saved above. Alternatively just use 
#     the table in memory (table2 <- taxonomic_info)
table2 <- read.csv(tbname, sep ="\t", header=T, as.is=TRUE) 
all_otus <- table2$qseqid
table2 <- table2[grepl("Streptophyta",table2$taxpath),] # retain Streptophyta
#discard not inventoried groups of Streptophyta
outgroups <- c("Chlorophyta","Sphagnopsida","Jungermanniopsida",
               "Bryopsida","Polytrichopsida","NA")
for (n in seq(1:length(outgroups))){
  table2 <- table2[!grepl(outgroups[n],table2$taxpath),]  
}
ingroup_otus <- table2$qseqid
outgroup_otus <- setdiff(all_otus,ingroup_otus)
tbname2 <- file.path(path,"Table_otu_taxonomy_plant_dbotu.txt")
ingr <- file.path(path,"ingroup_otus_RDS_dbotu")
outgr <- file.path(path,"outgroup_otus_RDS_dbotu")
#save table
{write.table(table2, tbname2, sep="\t",quote=FALSE, col.names = NA)}
{saveRDS(ingroup_otus,ingr)}
{saveRDS(outgroup_otus,outgr)}

#Split taxonomic string into levels for the OTU data
tab_name <- file.path(path,"Table_otu_taxonomy_plant_dbotu.txt")
otutaxonomy <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)
library(stringr)
otulevels <- str_split_fixed(otutaxonomy$taxpath, ";", 7)
otulevels <- gsub(".__","",otulevels)
otulevels <- as.data.frame(otulevels)
names(otulevels) <- c("kingdom","phylum","class","order","family","genus",
                      "species")
otutaxlevels <- cbind(otutaxonomy,otulevels)

tab_name <- file.path(path,"Table_otu_taxonomy_plant_levels_dbotu.txt")
{write.table(otutaxlevels, tab_name, sep="\t",quote=FALSE, col.names = NA)}

```
Now we have a table ("Table_otu_taxonomy.txt") with "full" taxonomic information for each plant-OTU. We also have a vector ("ingroup_otus") of ingroup (plant) OTUs to filter the OTU tables with. (The ingroup and outgroup vectors are also saved as RDS)  

###Filtering OTU tables to contain only ingroup taxa
Filter the OTU tables to only contain plants. Also keep only samples that do not represent negative controls, pcr controls, etc. 
```{r, eval=FALSE}
allFiles <- list.files(path)
allTabs <- allFiles[grepl("otutable$", allFiles)]
####allTabs <- allTabs[grepl("DADA2", allTabs)]
tab_names <- sort(as.vector(sapply(allTabs, function(x) strsplit(x, ".otutable")[[1]][1])))
read_tabs <- file.path(path, allTabs)
proc_tabs <- file.path(path, paste0(tab_names,".planttable"))
## Vector for filtering out controls, putting samples in right order, etc..
samples <- c("S001","S002","S003","S004","S005","S006","S007","S008","S067",
             "S009","S010","S011","S012","S013","S014","S040","S068","S015",
             "S016","S017","S018","S069","S070","S019","S020","S021","S022",
             "S024","S025","S026","S027","S041","S028","S029","S030","S032",
             "S033","S034","S035","S042","S036","S037","S038","S039","S086",
             "S087","S088","S089","S044","S071","S045","S046","S047","S048",
             "S049","S050","S051","S052","S053","S055","S056","S057","S058",
             "S090","S059","S060","S061","S062","S063","S064","S065","S066",
             "S072","S073","S074","S075","S076","S077","S078","S091","S079",
             "S080","S081","S082","S083","S084","S085","S092","S094","S095",
             "S096","S097","S098","S099","S100","S101","S102","S103","S104",
             "S106","S107","S108","S109","S133","S110","S111","S112","S113",
             "S114","S115","S116","S117","S118","S119","S120","S121","S122",
             "S123","S124","S134","S125","S126","S127","S129","S130","S131",
             "S132","S135","S136","S137") 
tab <- list()
keep_names <- list()
for(i in seq_along(read_tabs)) {
  tab[[i]] <- read.csv(read_tabs[i],sep='\t',header=T,as.is=TRUE,row.names = 1)
  ## setting rowname for SWARM tables where  name is in separate column
  if ("amplicon" %in% names(tab[[i]])) {  
    row.names(tab[[i]]) <- gsub(";size.*$","",tab[[i]]$amplicon)
  }
  seq_names <- row.names(tab[[i]])
  keep_names[[i]] <- seq_names %in% ingroup_otus
  # constrain table to contain only ingroup OTUs and sample columns
  tab[[i]] <- tab[[i]][keep_names[[i]],samples] 
  {write.table(tab[[i]], proc_tabs[i], sep="\t",quote=FALSE, col.names = NA)}
}
```
Now we have a new set of tables (with the suffix "planttable") containing only OTUs matching plant taxa. We now need to reextract the representative OTU sequences (centroids) for each ingroup-table to match those OTUs kept.  

For each of the OTU tables: Calculating the OTU richness plot wise. For each method/table calculate taxonomic redundancy, total OTU count, Number of unique taxonomic names, Number of taxonomic names which are also present in the observational data
```{r, eval=FALSE}
allFiles <- list.files(path)
all_plTabs <- allFiles[grepl("planttable$", allFiles)]
#all_prTabs <- allFiles[grepl("planttable.luluprocessed$", allFiles)]
all_Tabs <-  c(all_plTabs) #,all_prTabs)
read_tabs <- file.path(path, all_Tabs)
# Vector for filtering, etc. at this step redundant, but included for safety
samples <- c("S001","S002","S003","S004","S005","S006","S007","S008","S067",
             "S009","S010","S011","S012","S013","S014","S040","S068","S015",
             "S016","S017","S018","S069","S070","S019","S020","S021","S022",
             "S024","S025","S026","S027","S041","S028","S029","S030","S032",
             "S033","S034","S035","S042","S036","S037","S038","S039","S086",
             "S087","S088","S089","S044","S071","S045","S046","S047","S048",
             "S049","S050","S051","S052","S053","S055","S056","S057","S058",
             "S090","S059","S060","S061","S062","S063","S064","S065","S066",
             "S072","S073","S074","S075","S076","S077","S078","S091","S079",
             "S080","S081","S082","S083","S084","S085","S092","S094","S095",
             "S096","S097","S098","S099","S100","S101","S102","S103","S104",
             "S106","S107","S108","S109","S133","S110","S111","S112","S113",
             "S114","S115","S116","S117","S118","S119","S120","S121","S122",
             "S123","S124","S134","S125","S126","S127","S129","S130","S131",
             "S132","S135","S136","S137")  

tab_name <- file.path(path,"Table_otu_taxonomy_plant_levels_dbotu.txt")
otutaxonomy <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)

tab_name <- file.path(main_path,"Table_plants_2014_cleaned.txt")
Plant_data2014 <- read.table(tab_name, sep="\t", row.names = 1, header=TRUE,
                             as.is=TRUE)


Plant_richness <- colSums(Plant_data2014)
otu_richness <- data.frame(matrix(NA, nrow = 130, ncol = length(all_Tabs)))
names(otu_richness) <- all_Tabs
rel_redundancy <- vector()
total_otu <- vector()
mean_pident <- vector()
corcoeffs <- vector()
betadiversity <- vector()

##inserted
lm_intercept <- vector()
lm_slope <- vector()
read_sum <- vector()
Num_otu_taxa_method <- vector()
otu_taxa_method <- list()
singleton_share  <- vector()
doubleton_share  <- vector()
ab_diss <- list()
pa_diss <- list()
##inserted until here

for(i in seq_along(read_tabs)) {
  tab <- read.csv(read_tabs[i],sep='\t',header=T,as.is=TRUE,row.names = 1) #read table
  tab <- tab[,samples] # order samples
  otu_richness[,i] = colSums(tab>0) # calculate plot wise richness
  amp_index <- row.names(tab) #OTU id's of current table
  reftaxindex <- which(otutaxonomy$qseqid %in% amp_index) # index of which OTUs are present in the current table
  
  ##inserted
  perfect_match_index <- which(otutaxonomy$pident == 100 & otutaxonomy$qseqid %in% amp_index) # index of which OTUs are present in the current
  otu_taxa_method[[i]] <- names(table(otutaxonomy$species[perfect_match_index])) #Which species names have been identified in the current table
  Num_otu_taxa_method[i] <- length(otu_taxa_method[[i]]) # Number of plant species names
  ## until here
  
  mean_pident[[i]] <- mean(otutaxonomy$pident[reftaxindex]) # average genbank match %
  
  spec <- otutaxonomy$species[reftaxindex] # names of all OTUs
  redundancy <- sum((table(spec) -1)) # count of taxonomically redundant OTUs
  total_otu[i] <- nrow(tab)   #total number of OTUs present in the table
  betadiversity[i] <- total_otu[i]/mean(otu_richness[,i])
  rel_redundancy[i] <- redundancy/total_otu[i] #  relative redundancy
  # R^2 of linear regression of OTU richness vs plant richness
  corcoeffs[i] <- (cor(Plant_richness,otu_richness[,i]))^2
  lm_fit <- lm(otu_richness[,i]~ Plant_richness)
  lm_intercept[i] <- lm_fit$coefficients[1]
  lm_slope[i] <- lm_fit$coefficients[2]
  read_sum[i] <- sum(tab)
  
  #Inserted. community dissimilarity
  stable <- tab 
  trans_table <- t(stable)
  rowindex <- rowSums(trans_table) != 0
  trans_table <- trans_table[rowindex,]
  stand_table <- decostand(trans_table, "hellinger")
  ab_diss[[i]] <- vegdist(stand_table, method="bray", binary=FALSE)
  pa_diss[[i]] <- vegdist(stand_table, method="bray", binary=TRUE)
  #inserted until here
  
  #inserted
  tab2 <- tab
  tab2[tab2>1] <- 1
  singleton_share[i] <- sum(rowSums(tab2)==1)/total_otu[i]
  doubleton_share[i] <- sum(rowSums(tab2)==2)/total_otu[i]

}

p_table <- Plant_data2014
names(p_table) <- samples
trans_p_table <- t(p_table)
rowindex <- rowSums(trans_p_table) != 0
trans_p_table <- trans_p_table[rowindex,]
plant_pa_diss <- vegdist(trans_p_table, method="bray", binary=TRUE)


#MANTEL test for correlation with plant data on both presence absence (pa) tables and abundance tables (ab)
pa_vs_plant <- list() # Manteltest for plant data vd sequence data (pa)
ab_vs_plant <- list() # Manteltest for plant data vd sequence data (ab)
pa_vs_plant_statistic <- vector()  # Mantel statistic r (pa)
pa_vs_plant_signif <- vector() # significance level (pa)
ab_vs_plant_statistic <- vector() # Mantel statistic r (ab)
ab_vs_plant_signif <- vector() # significance level (ab)
 
for(i in 1:(length(read_tabs))) {
 pa_vs_plant[[i]] <- mantel(plant_pa_diss, pa_diss[[i]], method="pearson", permutations=999)
 pa_vs_plant_statistic[i] <- pa_vs_plant[[i]]$statistic
 pa_vs_plant_signif[i] <- pa_vs_plant[[i]]$signif
 ab_vs_plant[[i]] <- mantel(plant_pa_diss, ab_diss[[i]], method="pearson", permutations=999)
 ab_vs_plant_statistic[i] <- ab_vs_plant[[i]]$statistic
 ab_vs_plant_signif[i] <- ab_vs_plant[[i]]$signif
}

allFiles <- list.files(path)
all_plTabs <- allFiles[grepl("otutable$", allFiles)]
read_tabs <- file.path(path, all_plTabs)
tab_rc <- read.csv(read_tabs[1],sep='\t',header=T,as.is=TRUE,row.names = 1) #read table
sum(tab_rc) #6624089
tab_rc <- read.csv(read_tabs[2],sep='\t',header=T,as.is=TRUE,row.names = 1) #read table
sum(tab_rc) #6624089
#VALUES ADDED TO TABLES MANUALLY

#Synchronize names for methods, levels and curation state and 
#   collect table statistics in one table
method <- str_split_fixed(all_Tabs, "_", 3)[,1]
level <- str_split_fixed(all_Tabs, "_", 3)[,2]
level <- gsub(".planttable","",level)
level <- factor(level,levels = c("10", "0"))
#identify LULU curated tables

#Merge all results in one table
method_statistics <- data.frame(Method=method,Level=level,
                                Correlation=corcoeffs,
                                Redundancy=rel_redundancy,OTU_count=total_otu,
                                Mean_match=mean_pident,Beta=betadiversity,
                                Intercept = lm_intercept, Slope=lm_slope,
                                Total_readcount = read_sum, Taxa=Num_otu_taxa_method, 
                                Singleton=singleton_share,Doubleton=doubleton_share,
                                Com_dissim_PA_stat=pa_vs_plant_statistic,
                                Com_dissim_PA_sig=pa_vs_plant_signif,
                                Com_dissim_AB_stat=ab_vs_plant_statistic,
                                Com_dissim_AB_sig=ab_vs_plant_signif)

tab_name <- file.path(path,"Table_method_statistics_dbotu.txt")
{write.table(method_statistics, tab_name, sep="\t",quote=FALSE, col.names = NA)}



```

Construct a full plant richness vs OTU richness table and synchronize names for methods, levels and curation state
```{r,eval=FALSE}
# add Plant richness to OTU richness dataframe
richness_incl_obs <- cbind(Obs_richness=Plant_richness,otu_richness) 
total_richness_df <- gather(richness_incl_obs, key=Method, 
                            value=OTU_richness,-1)

method <- str_split_fixed(total_richness_df$Method, "_", 3)[,1]
level <- str_split_fixed(total_richness_df$Method, "_", 3)[,2]
level <- gsub(".planttable","",level)

level <- factor(level,levels = c("10", "0"))
total_richness_df2 <- data.frame(Method=method,Level=level,
                                 Obs=total_richness_df$Obs_richness,
                                 OTU=total_richness_df$OTU_richness)

#save a long formatted table for ggplot
tab_name <- file.path(path,"Table_richness_calculations_long_dbotu.txt")
{write.table(total_richness_df2, tab_name, sep="\t",quote=FALSE, col.names = NA)}

#save a wide formatted table for overview
tab_name <- file.path(path,"Table_richness_calculations_wide_dbotu.txt")
{write.table(richness_incl_obs, tab_name, sep="\t",quote=FALSE, col.names = NA)}


formula <- y ~ x
#Plot full x/y plots
dbotu_onestep_plot <- ggplot(total_richness_df2, aes(x=Obs,y=OTU)) +
  geom_point(pch=21,size=1, alpha = 0.8) +
  geom_abline(intercept = 0, linetype =2) +
  facet_grid(Method ~Level) +
  xlab("Plant richness") +
  ylab("OTU richness") +
  geom_smooth(method = "lm", se = F) +
  stat_poly_eq(geom = "label", 
               alpha = 0.5,aes(label = paste(..eq.label.., 
                                             ..rr.label.., sep = "~~~")), 
               formula = formula, label.x.npc = "left", 
               label.y.npc = "top", parse = TRUE, size = 3, label.size = NA) +
  scale_color_brewer(palette = "Set1") + theme_bw() + 
  theme(text = element_text(size=8))

namedbotu <- file.path(path,"dbotu_onestep_plotRDS")
saveRDS(dbotu_onestep_plot,"dbotu_onestep_plotRDS")

```
