---
title: "K - Benchmarking dbotu3 curation and singleton culling"
author: "Tobias G Frøslev"
date: "11/6/2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file records the post-clustering curation test for dbotu3, and effect of singleton culling for the manuscript "Reliable biodiversity metrics from co-occurence based post-clustering curation of amplicon data".  

For each table the following metrics are calculated:  
(a) Linear regression of OTU richness vs Plant richness for the 130 samples (including r^2^ value), 
(b) Number of OTUs,  
(c) Taxonomic redundancy,  
(d) Betadiversity (species/OTU turnover between sites), and  
(f) Distribution of best reference database matches

This step should be carried out after the LULU curation of the OTU tables documented in the file: E_Taxonomic_filtering.Rmd    
NB: All markdown chuncks are set to "eval=FALSE". Change these accordingly. Also code blocks to be run outside R, has been #'ed out. Change this accordingly.
  
###Analysis files  
This step is dependent on the presence of OTU tables (un-curated tables and the corresponding tables curated with LULU)  

Setting directories and libraries etc
```{r, eval=FALSE}
setwd("~/analyses")
main_path <- getwd()
path <- file.path(main_path, "otutables_processsing_dbotu")
library(stringr)
library(dplyr)
library(tidyr)
require(vegan)
```

Make 'singleton culled' versions of all unprocessed tables
```{r, eval=FALSE}
allFiles <- list.files(path)
all_plTabs <- allFiles[grepl("planttable$", allFiles)]
all_Tabs <-  c(all_plTabs)
read_tabs <- file.path(path, all_Tabs)
proc_tabs <- file.path(path, paste0(all_Tabs,"_xsingletons"))
# Vector for filtering, etc. at this step redundant, but included for safety
samples <- c("S001","S002","S003","S004","S005","S006","S007","S008","S067",
             "S009","S010","S011","S012","S013","S014","S040","S068","S015",
             "S016","S017","S018","S069","S070","S019","S020","S021","S022",
             "S024","S025","S026","S027","S041","S028","S029","S030","S032",
             "S033","S034","S035","S042","S036","S037","S038","S039","S086",
             "S087","S088","S089","S044","S071","S045","S046","S047","S048",
             "S049","S050","S051","S052","S053","S055","S056","S057","S058",
             "S090","S059","S060","S061","S062","S063","S064","S065","S066",
             "S072","S073","S074","S075","S076","S077","S078","S091","S079",
             "S080","S081","S082","S083","S084","S085","S092","S094","S095",
             "S096","S097","S098","S099","S100","S101","S102","S103","S104",
             "S106","S107","S108","S109","S133","S110","S111","S112","S113",
             "S114","S115","S116","S117","S118","S119","S120","S121","S122",
             "S123","S124","S134","S125","S126","S127","S129","S130","S131",
             "S132","S135","S136","S137")  


for(i in seq_along(read_tabs)) {
  tab <- read.csv(read_tabs[i],sep='\t',header=T,as.is=TRUE,row.names = 1)
  tab <- tab[,samples] # order samples
  tab[tab == 1] <- 0  ###REMOVE SINGLETONS
  tab <- tab[which(rowSums(tab) > 0),]
  {write.table(tab, proc_tabs[i], sep="\t",quote=FALSE, 
               col.names = NA)}
}
```

Apply dbotu3 as a post-clustering alternative to LULU
```{r}
#Copy all OTU tables to new directories for processing
# mkdir -p Processing_a0
# cp otutables_processing/*planttable dbotu3/Processing_a0/
# cp otutables_processing/*plantcentroids dbotu3/Processing_a0/
# cp otutables_processing/*planttable dbotu3/Processing_a10
# cp otutables_processing/*plantcentroids dbotu3/Processing_a10

#Copy all OTU tables to new directory for processing
# cd dbotu3/Processing_a0
# for f in *planttable; do
#  log=${f/planttable/log}
#  table=${f}
#  centroids=${f/planttable/plantcentroids}
#  output=${f/planttable/planttable_dbotuprocessed10}
#  python dbotu3tgf.py --dist 0.16 --abund 0 --log $log --output $output $table $centroids
# done

# cd dbotu3/Processing_a10
# for f in *planttable; do
#  log=${f/planttable/log}
#  table=${f}
#  centroids=${f/planttable/plantcentroids}
#  output=${f/planttable/planttable_dbotuprocessed10}
#  python dbotu3tgf.py --dist 0.16 --abund 10 --log $log --output $output $table $centroids
# done
```

For each of the OTU tables: Calculating biodiversity metrics (all tables need to be placed in same directory).  
```{r, eval=FALSE}
allFiles <- list.files(path)
all_plTabs <- allFiles[grepl("planttable$", allFiles)]
all_prTabs <- allFiles[grepl("planttable_luluprocessed$", allFiles)]
all_prTabs_xsingle <- allFiles[grepl("planttable_xsingletons$", allFiles)]
all_prTabs_dbotu <- allFiles[grepl("planttable_dbotuprocessed", allFiles)]

all_Tabs <-  c(all_plTabs,all_prTabs,all_prTabs_xsingle,all_prTabs_dbotu)
read_tabs <- file.path(path, all_Tabs)
# Vector for filtering, etc. at this step redundant, but included for safety
samples <- c("S001","S002","S003","S004","S005","S006","S007","S008","S067",
             "S009","S010","S011","S012","S013","S014","S040","S068","S015",
             "S016","S017","S018","S069","S070","S019","S020","S021","S022",
             "S024","S025","S026","S027","S041","S028","S029","S030","S032",
             "S033","S034","S035","S042","S036","S037","S038","S039","S086",
             "S087","S088","S089","S044","S071","S045","S046","S047","S048",
             "S049","S050","S051","S052","S053","S055","S056","S057","S058",
             "S090","S059","S060","S061","S062","S063","S064","S065","S066",
             "S072","S073","S074","S075","S076","S077","S078","S091","S079",
             "S080","S081","S082","S083","S084","S085","S092","S094","S095",
             "S096","S097","S098","S099","S100","S101","S102","S103","S104",
             "S106","S107","S108","S109","S133","S110","S111","S112","S113",
             "S114","S115","S116","S117","S118","S119","S120","S121","S122",
             "S123","S124","S134","S125","S126","S127","S129","S130","S131",
             "S132","S135","S136","S137")  

tab_name <- file.path(main_path,"Table_otu_taxonomy_plant_levels.txt")
otutaxonomy <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)

tab_name <- file.path(main_path,"Table_plants_2014_cleaned.txt")
Plant_data2014 <- read.table(tab_name, sep="\t", row.names = 1, header=TRUE,
                             as.is=TRUE)

Plant_richness <- colSums(Plant_data2014)
otu_richness <- data.frame(matrix(NA, nrow = 130, ncol = length(all_Tabs)))
names(otu_richness) <- all_Tabs
rel_redundancy <- vector()
total_otu <- vector()
mean_pident <- vector()
corcoeffs <- vector()
betadiversity <- vector()

##inserted
lm_intercept <- vector()
lm_slope <- vector()
read_sum <- vector()
Num_otu_taxa_method <- vector()
otu_taxa_method <- list()
singleton_share  <- vector()
doubleton_share  <- vector()
ab_diss <- list()
pa_diss <- list()
##inserted until here

for(i in seq_along(read_tabs)) {
  tab <- read.csv(read_tabs[i],sep='\t',header=T,as.is=TRUE,row.names = 1) #read table
  tab <- tab[,samples] # order samples
  otu_richness[,i] = colSums(tab>0) # calculate plot wise richness
  amp_index <- row.names(tab) #OTU id's of current table
  reftaxindex <- which(otutaxonomy$qseqid %in% amp_index) # index of which OTUs are present in the current table
  
  ##inserted
  perfect_match_index <- which(otutaxonomy$pident > 99.49 & otutaxonomy$qseqid %in% amp_index) # index of which OTUs are present in the current
  otu_taxa_method[[i]] <- names(table(otutaxonomy$species[perfect_match_index])) #Which species names have been identified in the current table
  Num_otu_taxa_method[i] <- length(otu_taxa_method[[i]]) # Number of plant species names
  ##inserted until here
  
  mean_pident[[i]] <- mean(otutaxonomy$pident[reftaxindex]) # average genbank match %
  
  spec <- otutaxonomy$species[reftaxindex] # names of all OTUs
  redundancy <- sum((table(spec) -1)) # count of taxonomically redundant OTUs
  total_otu[i] <- nrow(tab)   #total number of OTUs present in the table
  betadiversity[i] <- total_otu[i]/mean(otu_richness[,i])
  redundancy <- sum((table(spec) -1))
  rel_redundancy[i] <- redundancy/total_otu[i] #  relative redundancy
  # R^2 of linear regression of OTU richness vs plant richness
  corcoeffs[i] <- (cor(Plant_richness,otu_richness[,i]))^2
  lm_fit <- lm(otu_richness[,i]~ Plant_richness)
  lm_intercept[i] <- lm_fit$coefficients[1]
  lm_slope[i] <- lm_fit$coefficients[2]
  read_sum[i] <- sum(tab)
  
  #Inserted. community dissimilarity
  stable <- tab 
  trans_table <- t(stable)
  rowindex <- rowSums(trans_table) != 0
  trans_table <- trans_table[rowindex,]
  stand_table <- decostand(trans_table, "hellinger")
  ab_diss[[i]] <- vegdist(stand_table, method="bray", binary=FALSE)
  pa_diss[[i]] <- vegdist(stand_table, method="bray", binary=TRUE)
  #inserted until here
  
  #inserted
  tab2 <- tab
  tab2[tab2>1] <- 1
  singleton_share[i] <- sum(rowSums(tab2)==1)/total_otu[i]
  doubleton_share[i] <- sum(rowSums(tab2)==2)/total_otu[i]
}

#Synchronize names for methods, levels and curation state and 
#   collect table statistics in one table
method <- str_split_fixed(all_Tabs, "_", 3)[,1]
method[method == "DADA2"] <- "DADA2(+VS)"
method[method == "DADA2VSEARCH"] <- "DADA2(+VS)"
level <- str_split_fixed(all_Tabs, "_", 3)[,2]
level <- gsub(".planttable","",level)
level[level == "0.95"] <- "95"
level[level == "0.96"] <- "96"
level[level == "0.97"] <- "97"
level[level == "0.98"] <- "98"
level[level == "0.985"] <- "98.5"
level[level == "NO"] <- "99/100"
level[level == "3"] <- "99/100"
level[level == "5"] <- "98.5"
level[level == "7"] <- "98"
level[level == "10"] <- "97"
level[level == "13"] <- "96"
level[level == "15"] <- "95"
level <- factor(level,levels = c("99/100", "98.5", "98", "97", "96", "95"))

#identify curated tables
processed <- str_split_fixed(all_Tabs, "_", 3)[,3]
luluindex <- which(processed == "luluprocessed")
dbotu_a10_index <- which(processed == "dbotuprocessed10")
dbotu_a0_index <- which(processed == "dbotuprocessed0")
singleton_index <- which(processed == "xsingletons")

raw_index <- c(luluindex,dbotu_a10_index,dbotu_a0_index,singleton_index)

processed[dbotu_a10_index] <- "dbotu10"
processed[dbotu_a0_index] <- "dbotu0"
processed[singleton_index] <- "xsingle"
processed[luluindex] <- "curated"
processed[-raw_index] <- "raw"


#Merge all results in one table
method_statistics <- data.frame(Method=method,Level=level,
                                Curated=processed,Correlation=corcoeffs,
                                Redundancy=rel_redundancy,OTU_count=total_otu,
                                Mean_match=mean_pident,Beta=betadiversity,
                                Intercept = lm_intercept, Slope=lm_slope,
                                Total_readcount = read_sum, Taxa=Num_otu_taxa_method, 
                                Singleton=singleton_share,Doubleton=doubleton_share)

tab_name <- file.path(main_path,"Table_method_statistics_benchmarking.txt")
{write.table(method_statistics, tab_name, sep="\t",quote=FALSE, col.names = NA)}

```

Construct a full plant richness vs OTU richness table and synchronize names for methods, levels and curation state
```{r,eval=FALSE}
# add Plant richness to OTU richness dataframe
richness_incl_obs <- cbind(Obs_richness=Plant_richness,otu_richness) 
total_richness_df <- gather(richness_incl_obs, key=Method, 
                            value=OTU_richness,-1)

method <- str_split_fixed(total_richness_df$Method, "_", 3)[,1]
method[method == "DADA2"] <- "DADA2(+VS)"
method[method == "DADA2VSEARCH"] <- "DADA2(+VS)"
level <- str_split_fixed(total_richness_df$Method, "_", 3)[,2]
level <- gsub(".planttable","",level)
level[level == "0.95"] <- "95"
level[level == "0.96"] <- "96"
level[level == "0.97"] <- "97"
level[level == "0.98"] <- "98"
level[level == "0.985"] <- "98.5"
level[level == "NO"] <- "99/100"
level[level == "3"] <- "99/100"
level[level == "5"] <- "98.5"
level[level == "7"] <- "98"
level[level == "10"] <- "97"
level[level == "13"] <- "96"
level[level == "15"] <- "95"
level[level == "100"] <- "99"
level <- factor(level,levels = c("99/100", "98.5", "98", "97", "96", "95"))
processed <- str_split_fixed(total_richness_df$Method, "_", 3)[,3]

luluindex <- which(processed == "luluprocessed")
dbotu_a10_index <- which(processed == "dbotuprocessed10")
dbotu_a0_index <- which(processed == "dbotuprocessed0")
singleton_index <- which(processed == "xsingletons")

raw_index <- c(luluindex,dbotu_a10_index,dbotu_a0_index,singleton_index)
#raw_index <- c(luluindex,dbotu_a10_index)

processed[dbotu_a10_index] <- "dbotu10"
processed[dbotu_a0_index] <- "dbotu0"
processed[singleton_index] <- "xsingle"
processed[luluindex] <- "curated"
processed[-raw_index] <- "raw"

total_richness_df2 <- data.frame(Method=method,Level=level,Curated=processed,
                                 Obs=total_richness_df$Obs_richness,
                                 OTU=total_richness_df$OTU_richness)

#save a long formatted table for ggplot
tab_name <- file.path(main_path,"Table_richness_calculations_long_bdotu3_benchmarking.txt")
{write.table(total_richness_df2, tab_name, sep="\t",quote=FALSE, col.names = NA)}

#save a wide formatted table for overview
tab_name <- file.path(main_path,"Table_richness_calculations_wide_bdotu3_benchmarking.txt")
{write.table(richness_incl_obs, tab_name, sep="\t",quote=FALSE, col.names = NA)}
```

Construct a table with the best match (%) for each OTU pr method/table separating retained and discarded OTUs.
```{r,eval=FALSE}
setwd("~/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses")
main_path <- getwd()
path <- file.path(main_path, "otutables_processsing_dbotu")

allFiles <- list.files(path)
all_plTabs <- allFiles[grepl("planttable$", allFiles)]
all_prTabs_dbotu0 <- allFiles[grepl("planttable_dbotuprocessed0", allFiles)]
all_prTabs_dbotu10 <- allFiles[grepl("planttable_dbotuprocessed10", allFiles)]
all_prTabs_xsingle <- allFiles[grepl("planttable_xsingletons", allFiles)]

read_tabs_raw <- file.path(path, all_plTabs)
read_tabs_dbotu3a0 <- file.path(path, all_prTabs_dbotu0)
read_tabs_dbotu3a10 <- file.path(path, all_prTabs_dbotu10)
read_tabs_xsingle <- file.path(path, all_prTabs_xsingle)

tab_names <- sort(as.vector(sapply(all_plTabs, function(x) strsplit(x, ".planttable")[[1]][1])))

tab_name <- file.path(main_path,"Table_otu_taxonomy_plant_levels.txt")
otutaxonomy <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)

retained_avg <- vector()
discarded_avg <- vector()
pident_frame_a0 <- data.frame()
pident_frame_a10 <- data.frame()
pident_frame_xs <- data.frame()
for(i in seq_along(read_tabs_raw)) {
  tab_raw <- read.csv(read_tabs_raw[i],sep='\t',header=T,as.is=TRUE,row.names = 1) #read table
  tab_a0 <- read.csv(read_tabs_dbotu3a0[i],sep='\t',header=T,as.is=TRUE,row.names = 1) #read table
  tab_a10 <- read.csv(read_tabs_dbotu3a10[i],sep='\t',header=T,as.is=TRUE,row.names = 1) #read table
  tab_xs <- read.csv(read_tabs_xsingle[i],sep='\t',header=T,as.is=TRUE,row.names = 1) #read table

  retained_amplicons_a0 <- row.names(tab_a0)
  retained_amplicons_a10 <- row.names(tab_a10)
  retained_amplicons_xs <- row.names(tab_xs)
  
  discarded_amplicons_a0 <- setdiff(row.names(tab_raw),retained_amplicons_a0)
  discarded_amplicons_a10 <- setdiff(row.names(tab_raw),retained_amplicons_a0)
  discarded_amplicons_xs <- setdiff(row.names(tab_raw),retained_amplicons_xs)

  number_observations_a0 <- length(retained_amplicons_a0)+length(discarded_amplicons_a0)
  number_observations_a10 <- length(retained_amplicons_a10)+length(discarded_amplicons_a10)
  number_observations_xs <- length(retained_amplicons_xs)+length(discarded_amplicons_xs)

  retained_index_a0 <- which(otutaxonomy$qseqid %in% retained_amplicons_a0)
  retained_index_a10 <- which(otutaxonomy$qseqid %in% retained_amplicons_a10)
  retained_index_xs <- which(otutaxonomy$qseqid %in% retained_amplicons_xs)

  discarded_index_a0 <- which(otutaxonomy$qseqid %in% discarded_amplicons_a0)
  discarded_index_a10 <- which(otutaxonomy$qseqid %in% discarded_amplicons_a10)
  discarded_index_xs <- which(otutaxonomy$qseqid %in% discarded_amplicons_xs)
  
  retained_pident_a0 <- otutaxonomy$pident[retained_index_a0]
  retained_pident_a10 <- otutaxonomy$pident[retained_index_a10]
  retained_pident_xs <- otutaxonomy$pident[retained_index_xs]
  
  discarded_pident_a0 <- otutaxonomy$pident[discarded_index_a0]
  discarded_pident_a10 <- otutaxonomy$pident[discarded_index_a10]
  discarded_pident_xs <- otutaxonomy$pident[discarded_index_xs]
  
  method_pident_a0 <- rep(tab_names[i],number_observations_a0)
  method_pident_a10 <- rep(tab_names[i],number_observations_a10)
  method_pident_xs <- rep(tab_names[i],number_observations_xs)
  
  retained_or_discarded_a0 <- c(rep("retained",length(retained_amplicons_a0)),
                             rep("discarded",length(discarded_amplicons_a0)))
  retained_or_discarded_a10 <- c(rep("retained",length(retained_amplicons_a10)),
                             rep("discarded",length(discarded_amplicons_a10)))
  retained_or_discarded_xs <- c(rep("retained",length(retained_amplicons_xs)),
                             rep("discarded",length(discarded_amplicons_xs)))
  
  pident_a0 <- c(retained_pident_a0,discarded_pident_a0)
  pident_a10 <- c(retained_pident_a10,discarded_pident_a10)
  pident_xs <- c(retained_pident_xs,discarded_pident_xs)

  current_pident_frame_a0 <- data.frame(method_pident_a0,retained_or_discarded_a0,pident_a0)
  current_pident_frame_a10 <- data.frame(method_pident_a10,retained_or_discarded_a10,pident_a10)
  current_pident_frame_xs <- data.frame(method_pident_xs,retained_or_discarded_xs,pident_xs)
  
  pident_frame_a0 <- rbind(pident_frame_a0,current_pident_frame_a0)
  pident_frame_a10 <- rbind(pident_frame_a10,current_pident_frame_a10)
  pident_frame_xs <- rbind(pident_frame_xs,current_pident_frame_xs)
}

names(pident_frame_a0) <- c("method_pident","retained_or_discarded","pident")
names(pident_frame_a10) <- c("method_pident","retained_or_discarded","pident")
names(pident_frame_xs) <- c("method_pident","retained_or_discarded","pident")


pident_frame_a0$pca <- "dbotu a0"
pident_frame_a10$pca <- "dbotu a10"
pident_frame_xs$pca <- "xsingle"

pident_frame <- rbind(pident_frame_a0,pident_frame_a10,pident_frame_xs)

#Synchronize names for methods, levels and curation state
pident_frame$level_pident <- 
 str_split_fixed(as.character(pident_frame$method_pident), "_", 2)[,2]
pident_frame$method_pident <- 
 str_split_fixed(as.character(pident_frame$method_pident), "_", 2)[,1]
pident_frame$method_pident[pident_frame$method_pident == "DADA2"] <- 
 "DADA2(+VS)"
pident_frame$method_pident[pident_frame$method_pident == "DADA2VSEARCH"] <- 
 "DADA2(+VS)"
pident_frame$level_pident[pident_frame$level_pident == "0.95"] <- "95"
pident_frame$level_pident[pident_frame$level_pident == "0.96"] <- "96"
pident_frame$level_pident[pident_frame$level_pident == "0.97"] <- "97"
pident_frame$level_pident[pident_frame$level_pident == "0.98"] <- "98"
pident_frame$level_pident[pident_frame$level_pident == "0.985"] <- "98.5"
pident_frame$level_pident[pident_frame$level_pident == "NO"] <- "99/100"
pident_frame$level_pident[pident_frame$level_pident == "3"] <- "99/100"
pident_frame$level_pident[pident_frame$level_pident == "5"] <- "98.5"
pident_frame$level_pident[pident_frame$level_pident == "7"] <- "98"
pident_frame$level_pident[pident_frame$level_pident == "10"] <- "97"
pident_frame$level_pident[pident_frame$level_pident == "13"] <- "96"
pident_frame$level_pident[pident_frame$level_pident == "15"] <- "95"
pident_frame$level_pident <- 
 factor(pident_frame$level_pident,levels = c("99/100", "98.5", 
                                             "98", "97", "96", "95"))

#Combine with results from LULU
tab_name <- file.path(main_path,"Table_OTU_match_rates.txt")
pident_lulu <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)

pident_lulu$pca <- "LULU"

pident_frame <- pident_frame[,c(1,2,3,5,4)]
pident_lulu <- pident_lulu[,c(2,3,4,5,6)]
pident_frame <- rbind(pident_lulu,pident_frame)

pident_frame$pca <- 
 factor(pident_frame$pca,levels = c("LULU", "dbotu a0","dbotu a10","xsingle"))

tab_name <- file.path(main_path,"Table_OTU_match_rates_benchmarking.txt")
{write.table(pident_frame, tab_name, sep="\t",quote=FALSE, col.names = NA)}

```
