---
output:
  word_document: default
  pdf_document: default
  html_document: default
---
```{r,echo=FALSE, message=FALSE}
setwd("~/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses")
main_path <- getwd() 
path <- file.path(main_path, "otutables_processing")
library(stringr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpmisc)
library(cowplot)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

**OTU richness vs. plant richness - LULU curation, singleton culling, curation with dbotu3**  
```{r dpi=300,echo=FALSE}
tab_name <- file.path(main_path,"Table_richness_calculations_long_bdotu3_benchmarking.txt")
total_richness_df2 <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)
total_richness_df2$Level <- 
 factor(total_richness_df2$Level,levels = c("99/100", "98.5", "98", "97", 
                                            "96", "95"))
total_richness_df2$Method <- 
 factor(total_richness_df2$Method,levels = c("VSEARCH", "SWARM", "DADA2(+VS)", 
                                             "CROP"))

total_richness_df2$Curated[total_richness_df2$Curated == "curated"] <- "LULU"

total_richness_df2$Curated <- 
 factor(total_richness_df2$Curated,levels = c("raw","xsingle","dbotu10","dbotu0","LULU"))

formula <- y ~ x

#Plot full x/y plots
ggplot(total_richness_df2, aes(x= Obs, y= OTU, color = Curated)) +
  geom_point(pch=21,size=1, alpha = 0.8) +
  geom_abline(intercept = 0, linetype =2) +
  facet_grid(Method ~Level) +
  xlab("Plant richness") +
  ylab("OTU richness") +
  geom_smooth(method = "lm", se = F,size=0.5) +
  #stat_poly_eq(geom = "label", 
  #             alpha = 0.5,aes(label = paste(..eq.label.., 
  #                                           ..rr.label.., sep = "~~~")), 
  #             formula = formula, label.x.npc = "left", 
  #             label.y.npc = "top", parse = TRUE, size = 1, label.size = NA) +
  scale_fill_manual(values=cbPalette) + theme_bw() + 
  theme(text = element_text(size=8)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

**Figure 6a.** OTU richness (number of OTUs in each soil sample from the 130 sites) is plotted on the y-axis. Plant richness (number of plant species observed in each of the 130 40m x 40m sites) is plotted on the x-axis. Values are shown for un-curated OTU tables, tables with singletons removed (xsingle), tables curated with dbotu3, abundance criterion 10 (dbotu10), tables curated with dbotu3, abundance criterion 0 (dbotu0) and tables curated with LULU. The dashed line is an identity to evaluate whether the OTU count overestimates (to the left of the line) or underestimates (to the right) the plant richness. The “99/100%” clustering level denotes the pure DADA2 approach (100%) and SWARM with a d-value of 3 (99%). (a) shows the full y-axis, whereas (b) shows a truncated y-axis to better illustrate the correlations, by excluding the extreme values. Statistics of the regression can be seen in Supplementary Table 3. Although difficult to see from the plots, correspondence with plant data was best for the LULU curation.   

\pagebreak

**OTU richness vs. plant richness - LULU curation, singleton culling, curation with dbotu3**  
```{r dpi=300,echo=FALSE}
ggplot(total_richness_df2, aes(x= Obs, y= OTU, col = Curated)) +
  geom_point(pch=21,size=1, alpha = 0.8) +
  coord_cartesian(ylim = c(0, 200)) +
  geom_abline(intercept = 0, linetype =2) +
  facet_grid(Method ~Level) +
  xlab("Plant richness") +
  ylab("OTU richness") +
  geom_smooth(method = "lm", se = F, size=0.5) +
  #stat_poly_eq(geom = "label", 
  #             alpha = 0.7,aes(label = paste(..eq.label.., 
  #                                           ..rr.label.., sep = "~~~")), 
  #             formula = formula, label.x.npc = "left", 
  #             label.y.npc = 0.2, label.size = NA, parse = TRUE, size = 1) +
  scale_fill_manual(values=cbPalette) + 
  theme_bw() + 
  theme(text = element_text(size=8)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

**Figure 6b (truncated y-axis).** OTU richness (number of OTUs in each soil sample from the 130 sites) is plotted on the y-axis. Plant richness (number of plant species observed in each of the 130 40m x 40m sites) is plotted on the x-axis. Values are shown for un-curated OTU tables, tables with singletons removed (xsingle), tables curated with dbotu3, abundance criterion 10 (dbotu10), tables curated with dbotu3, abundance criterion 0 (dbotu0) and tables curated with LULU. The dashed line is an identity to evaluate whether the OTU count overestimates (to the left of the line) or underestimates (to the right) the plant richness. The “99/100%” clustering level denotes the pure DADA2 approach (100%) and SWARM with a d-value of 3 (99%). (a) shows the full y-axis, whereas (b) shows a truncated y-axis to better illustrate the correlations, by excluding the extreme values. Statistics of the regression can be seen in Supplementary Table 3. Although difficult to see from the plots, correspondence with plant data was best for the LULU curation.      

\pagebreak

```{r dpi=300,echo=FALSE}
tab_name <- file.path(main_path,"Table_method_statistics_benchmarking.txt")
method_statistics <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)
method_statistics$Level <- 
 factor(method_statistics$Level,levels = c("99/100", "98.5", "98", "97", 
                                           "96", "95"))
method_statistics$Method <- 
 factor(method_statistics$Method,levels = c("VSEARCH", "SWARM", "DADA2(+VS)", 
                                            "CROP"))
method_statistics$Curated[which(method_statistics$Curated == "curated")] <- "LULU"

method_statistics$Curated <- 
 factor(method_statistics$Curated,levels = c("raw","xsingle","dbotu10","dbotu0","LULU"))
```


**Total number of OTUs - LULU curation, singleton culling, curation with dbotu3** 
```{r dpi=300,echo=FALSE}
ggplot(method_statistics,aes(x=Curated,weights=OTU_count,fill=Curated)) +
  geom_bar(position="dodge") + 
  geom_hline(yintercept = 564,linetype = 2) +
  facet_grid(Method ~Level) +
  xlab("") +
  ylab("OTUs") +
  scale_fill_manual(values=cbPalette) +
  theme_bw()+
  theme(text = element_text(size=8)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

**Figure 7a. (fixed y-axis)** Total method level OTU richness is plotted on the y-axis. Values are shown for un-curated OTU tables, tables with singletons removed (xsingle), tables curated with dbotu3, abundance criterion 10 (dbotu10), tables curated with dbotu3, abundance criterion 0 (dbotu0) and tables curated with LULU. The dashed line indicates the total number of species (564) observed in the study for comparison. (a) shows a fixed y-axis to compare across methods, (b) shows a flexible y-axis for better comparison of the low richness CROP method. LULU consistently performed better on tables from the two greedy algorithms, but dbotu3 resulted in a comparable curation for the other approaches.  

\pagebreak

**Total number of OTUs - LULU curation, singleton culling, curation with dbotu3** 
```{r dpi=300,echo=FALSE}
ggplot(method_statistics,aes(x=Curated,weights=OTU_count,fill=Curated)) +
  geom_bar(position="dodge") + 
  geom_hline(yintercept = 564,linetype = 2) +
  facet_grid(Method ~Level,scale = "free_y") +
  xlab("") +
  ylab("OTUs") +
  scale_fill_manual(values=cbPalette) +
  theme_bw()+
  theme(text = element_text(size=8)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

**Figure 7b. (flexible y-axis)** Total method level OTU richness is plotted on the y-axis. Values are shown for un-curated OTU tables, tables with singletons removed (xsingle), tables curated with dbotu3, abundance criterion 10 (dbotu10), tables curated with dbotu3, abundance criterion 0 (dbotu0) and tables curated with LULU. The dashed line indicates the total number of species (564) observed in the study for comparison. The “99/100%” clustering level denotes the pure DADA2 approach (100%) and SWARM with a d-value of 3 (99%). (a) shows a fixed y-axis to compare across methods, (b) shows a flexible y-axis for better comparison of the low richness CROP method. LULU consistently performed better on tables from the two greedy algorithms, but dbotu3 resulted in a comparable curation for the other approaches.    

\pagebreak

**Taxonomic redundancy - LULU curation, singleton culling, curation with dbotu3**  
```{r dpi=300,echo=FALSE}
ggplot(method_statistics,aes(x=Curated,weights=Redundancy,fill=Curated)) +
  geom_bar(position="dodge") + 
  facet_grid(Method ~Level) +
  xlab("") +
  ylab("Taxonomic redundancy (percentage of OTUs with redundant 
       taxonomic annotation)") +
  scale_fill_manual(values=cbPalette) +
  theme_bw()+
  scale_y_continuous(labels = scales::percent) + 
  theme(text = element_text(size=8)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

**Figure 8.** Taxonomic redundancy (the proportion of OTUs with a redundant taxonomic assignment) is plotted on the y-axis. Values are shown for un-curated OTU tables, tables with singletons removed (xsingle), tables curated with dbotu3, abundance criterion 10 (dbotu10), tables curated with dbotu3, abundance criterion 0 (dbotu0) and tables curated with LULU. The “99/100%” clustering level denotes the pure DADA2 approach (100%) and SWARM with a d-value of 3 (99%). LULU performed best in all comparisons.   

\pagebreak

**Betadiversity - LULU curation, singleton culling, curation with dbotu3**  
```{r dpi=300,echo=FALSE}
#Get beta diversity of plant survey
tab_name <- file.path(main_path,"Table_plants_2014_cleaned.txt")
Plant_data2014 <- read.table(tab_name, sep="\t", row.names = 1, 
                             header=TRUE, as.is=TRUE)
Plant_richness <- colSums(Plant_data2014)
obs_beta <- nrow(Plant_data2014)/mean(Plant_richness)

ggplot(method_statistics,aes(x=Curated,weights=Beta,fill=Curated)) +
  geom_bar(position="dodge") + 
  geom_hline(yintercept = obs_beta,linetype = 2) +
  facet_grid(Method ~Level) +
  xlab("") +
  ylab("Betadiversity (total richness / avg. plot richness)") +
  scale_fill_manual(values=cbPalette) +
  theme_bw()+ theme(text = element_text(size=8)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

**Figure 9.** Betadiversity (calculated as total number of OTUs divided by the mean number of OTUs in the 130 sites) is plotted on the y-axis. Values are shown for un-curated OTU tables, tables with singletons removed (xsingle), tables curated with dbotu3, abundance criterion 10 (dbotu10), tables curated with dbotu3, abundance criterion 0 (dbotu0) and tables curated with LULU. The dashed line indicates the betadiversity of the plant data (17.23) observed in the study for comparison, calculated in the same way. The “99/100%” clustering level denotes the pure DADA2 approach (100%) and SWARM with a d-value of 3 (99%). All approaches result in more realistic betadiversity estimates, but LULU and dbotu3 (a=0) consistently performed best.   

\pagebreak

**Distribution of best matches of retained OTUs - LULU curation, singleton culling, curation with dbotu3** 
```{r dpi=300,echo=FALSE}
tab_name <- file.path(main_path,"Table_OTU_match_rates_benchmarking.txt")
pident_frame <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)

pident_frame$level_pident <- 
 factor(pident_frame$level_pident,levels = c("99/100", "98.5", "98", "97", 
                                             "96", "95"))
pident_frame$method_pident <- 
 factor(pident_frame$method_pident,levels = c("VSEARCH", "SWARM", "DADA2(+VS)", 
                                              "CROP"))

pident_frame$pca[pident_frame$pca == "dbotu a10"] <- "dbotu10"
pident_frame$pca[pident_frame$pca == "dbotu a0"] <- "dbotu0"

pident_frame$pca <- 
 factor(pident_frame$pca,levels = c("xsingle","dbotu10","dbotu0","LULU"))



pident_frameR <- filter(pident_frame, retained_or_discarded == "retained")
pident_frameD <- filter(pident_frame, retained_or_discarded == "discarded")

#Violin plot of distribution of best matches, all methods
ggplot(pident_frameR, aes(x=pca, 
                         y=pident,fill=pca)) +
  geom_violin() +
  facet_grid(method_pident~level_pident)+
  xlab("number of OTUs") +
  ylab("Best match on GenBank") +
  scale_fill_manual(values=cbPalette) +
  theme_bw()+
  theme(text = element_text(size=8)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

**Figure 10.** Density distribution of the best reference database match for all retained OTUs (percent identity (%) of best matching reference sequence on GenBank) is plotted as a violin plot. Values are shown for OTU tables with singletons removed, tables processed with dbotu3 (a=10), and tables processed with dbotu3 (a=10) and tables curated with LULU. The “99/100%” clustering level denotes the pure DADA2 approach (100%) and SWARM with a d-value of 3 (99%). Although distributions are remarkably similar, some of the distribution show that LULU retains a higher proportion of perfect matches.  

\pagebreak

**Distribution of best matches of discarded OTUs - LULU curation, singleton culling, curation with dbotu3**  
```{r dpi=300,echo=FALSE}
#Violin plot of distribution of best matches, all methods
ggplot(pident_frameD, aes(x=pca, 
                         y=pident,fill=pca)) +
  geom_violin() +
  facet_grid(method_pident~level_pident)+
  xlab("number of OTUs") +
  ylab("Best match on GenBank") +
  scale_fill_manual(values=cbPalette) +
  theme_bw()+
  theme(text = element_text(size=8)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

**Figure 11.** Density distribution of the best reference database match for all discarded OTUs (percent identity (%) of best matching reference sequence on GenBank) is plotted as a violin plot. Values are shown for OTU tables with singletons removed, tables processed with dbotu3 (a=10), and tables processed with dbotu3 (a=10) and tables curated with LULU. The “99/100%” clustering level denotes the pure DADA2 approach (100%) and SWARM with a d-value of 3 (99%). Distributions are very similar, and no clear differences can be seen.  

\pagebreak  
**OTU richness vs. plant richness - dbotu3 as a 'one-step' clustering tool**  
```{r dpi=300,echo=FALSE}
#get dbotu3 one-step plot
nameRDS <- file.path(main_path,"dbotu_onestep_plotRDS")
dbotu3_onestep_plot <- readRDS(nameRDS)
dbotu3_onestep_plot

```
**Figure 12.** OTU richness (number of OTUs in each soil sample from the 130 sites) is plotted on the y-axis. Plant richness (number of plant species observed in each of the 130 40m x 40m sites) is plotted on the x-axis. Values are shown for the two OTU tables produced with the dbotu3 algorithm as a one-step tool with two different abundance cutoff settings: abundance criterion 10 (intended for removing sequencing errors) and 0 (aimed at accounting for only sequencing error). The dashed line is an identity to evaluate whether the OTU count overestimates (to the left of the line) or underestimates (to the right) the plant richness. Statistics of the regression can be seen in the top.  


