---
title: "F - Plot wise curation statistics"
author: "Tobias G Frøslev"
date: "8/8/2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file records the calculation of plotwise taxonomic composition for the manuscript "Reliable biodiversity metrics from post clustering curation of amplicon data".  

This step should be carried out after the LULU curation of the OTU tables documented in the file: E_Taxonomic_filtering.Rmd    
NB: All markdown chuncks are set to "eval=FALSE". Change these accordingly. Also code blocks to be run outside R, has been #'ed out. Change this accordingly.

### Bioinformatic tools necessary
Make sure that you have the following bioinformatic tools in your PATH  
R packages: stringr  
  
###Analysis files  
This step is dependent on the presence of OTU tables (un-curated tables and the corresponding tables curated with LULU)  

Setting directories and libraries etc
```{r, eval=FALSE}
setwd("~/analyses")
setwd("~/Documents/BIOWIDE/BIOWIDE_MANUSCRIPTS/Alfa_diversity/analyses")
main_path <- getwd()
path <- file.path(main_path, "otutables_processing")
library(stringr)
```


```{r, eval=FALSE}
allFiles <- list.files(path)
all_plTabs <- allFiles[grepl("planttable$", allFiles)]
all_prTabs <- allFiles[grepl("planttable.luluprocessed$", allFiles)]
all_Tabs <-  c(all_plTabs,all_prTabs)
read_tabs <- file.path(path, all_Tabs)
# Vector for filtering, etc. at this step redundant, but included for safety
samples <- c("S001","S002","S003","S004","S005","S006","S007","S008","S067",
             "S009","S010","S011","S012","S013","S014","S040","S068","S015",
             "S016","S017","S018","S069","S070","S019","S020","S021","S022",
             "S024","S025","S026","S027","S041","S028","S029","S030","S032",
             "S033","S034","S035","S042","S036","S037","S038","S039","S086",
             "S087","S088","S089","S044","S071","S045","S046","S047","S048",
             "S049","S050","S051","S052","S053","S055","S056","S057","S058",
             "S090","S059","S060","S061","S062","S063","S064","S065","S066",
             "S072","S073","S074","S075","S076","S077","S078","S091","S079",
             "S080","S081","S082","S083","S084","S085","S092","S094","S095",
             "S096","S097","S098","S099","S100","S101","S102","S103","S104",
             "S106","S107","S108","S109","S133","S110","S111","S112","S113",
             "S114","S115","S116","S117","S118","S119","S120","S121","S122",
             "S123","S124","S134","S125","S126","S127","S129","S130","S131",
             "S132","S135","S136","S137")  

tab_name <- file.path(main_path,"Table_otu_taxonomy_plant_levels.txt")
otutaxonomy <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)

tab_name <- file.path(main_path,"Table_plants_2016_alfa.txt") 
Plant_data2016 <- read.table(tab_name, sep="\t", row.names = 1, header=TRUE, as.is=TRUE)

plant_obs_species <- list()
for (i in 1:130){
 plant_obs_species[[i]] <- rownames(Plant_data2016)[which(Plant_data2016[,i] == 1)]
}

otu_taxa_method <- list()

no_OTUsUU <- data.frame(matrix(NA, ncol = 130,  nrow= length(all_Tabs)))
perfectmatchesUU <- data.frame(matrix(NA, ncol = 130,  nrow= length(all_Tabs)))
refindsUU <- data.frame(matrix(NA, ncol = 130,  nrow= length(all_Tabs)))
unknownUU <- data.frame(matrix(NA, ncol = 130,  nrow= length(all_Tabs)))
redundantUU <- data.frame(matrix(NA, ncol = 130,  nrow= length(all_Tabs)))
imperfectUU <- data.frame(matrix(NA, ncol = 130,  nrow= length(all_Tabs)))

refound_species <- list()

  for(i in seq_along(read_tabs)) {
   tab <- read.csv(read_tabs[i],sep='\t',header=T,as.is=TRUE,row.names = 1) #read table
   tab <- tab[,samples] # order samples

   otu_taxa_methodX <- list()
   Num_otu_taxa_methodX <- rep(0, 130)
   refound_species_site <- list()
   for (io in 1:130){
    amp_index <- row.names(tab)[which(tab[,io] > 0)] #OTU id's of current table
    
    no_OTUsUU[i,io] <- length(amp_index)
    
    reftaxindex <- which(otutaxonomy$qseqid %in% amp_index) # index of which OTUs are present in the current table
    perfect_match_index <- which(otutaxonomy$pident == 100 & otutaxonomy$qseqid %in% amp_index) # index of which OTUs are present in the current
    perfectmatchesUU[i,io] <- length(perfect_match_index)
    if (length(perfect_match_index) > 0) {
     otu_taxa_methodX[[io]] <- names(table(otutaxonomy$species[perfect_match_index])) #Which species names have been identified in the current table
     if (length(otu_taxa_methodX[[io]]) > 0) {
      Num_otu_taxa_methodX[io] <- length(otu_taxa_methodX[[io]])
      refindsUU[i,io] <- sum(otu_taxa_methodX[[io]] %in% plant_obs_species[[io]])
      refound_species_site[[io]] <- otu_taxa_methodX[[io]][otu_taxa_methodX[[io]] %in% plant_obs_species[[io]]]
       unknownUU[i,io] <- sum(!otu_taxa_methodX[[io]] %in% plant_obs_species[[io]])
      redundantUU[i,io] <- perfectmatchesUU[i,io] - (refindsUU[i,io]+unknownUU[i,io])
      imperfectUU[i,io] <- no_OTUsUU[i,io] - perfectmatchesUU[i,io]
     }
    }
    refound_species[[i]] <- refound_species_site
   }
  }
  
#Calculating average plot wise curation effects
perfect_match_share <- perfectmatchesUU/no_OTUsUU
perfect_match_share_sd <- apply(perfect_match_share,1,sd,na.rm=TRUE)
perfect_match_share_mean <- apply(perfect_match_share,1,mean,na.rm=TRUE)

imperfect_share <- imperfectUU/no_OTUsUU
imperfect_share_sd <- apply(imperfect_share,1,sd,na.rm=TRUE)
imperfect_share_mean <- apply(imperfect_share,1,mean,na.rm=TRUE)

refinds_share <- refindsUU/no_OTUsUU
refinds_share_sd <- apply(refinds_share,1,sd,na.rm=TRUE)
refinds_share_mean <- apply(refinds_share,1,mean,na.rm=TRUE)

unknown_share <- unknownUU/no_OTUsUU
unknown_share_sd <- apply(unknown_share,1,sd,na.rm=TRUE)
unknown_share_mean <- apply(unknown_share,1,mean,na.rm=TRUE)

redundant_share <- redundantUU/no_OTUsUU
redundant_share_sd <- apply(redundant_share,1,sd,na.rm=TRUE)
redundant_share_mean <- apply(redundant_share,1,mean,na.rm=TRUE)

refinds_lost <- (refindsUU[1:20,]-refindsUU[21:40,])/refindsUU[1:20,]
refinds_lost[is.na(refinds_lost)] <- 0
refinds_lost_sd <- apply(refinds_lost,1,sd,na.rm=TRUE)
refinds_lost_mean <- apply(refinds_lost,1,mean,na.rm=TRUE)

method <- str_split_fixed(all_Tabs, "_", 3)[,1]
method[method == "DADA2"] <- "DADA2(+VS)"
method[method == "DADA2VSEARCH"] <- "DADA2(+VS)"
level <- str_split_fixed(all_Tabs, "_", 3)[,2]
level <- gsub(".planttable","",level)
level[level == "0.95"] <- "95"
level[level == "0.96"] <- "96"
level[level == "0.97"] <- "97"
level[level == "0.98"] <- "98"
level[level == "0.985"] <- "98.5"
level[level == "NO"] <- "99/100"
level[level == "3"] <- "99/100"
level[level == "5"] <- "98.5"
level[level == "7"] <- "98"
level[level == "10"] <- "97"
level[level == "13"] <- "96"
level[level == "15"] <- "95"
level <- factor(level,levels = c("99/100", "98.5", "98", "97", "96", "95"))

#identify LULU curated tables
processed <- str_split_fixed(all_Tabs, "_", 3)[,3]
luluindex <- which(processed == "luluprocessed")
processed[luluindex] <- "curated"
processed[-luluindex] <- "raw"

#Merge all results in one table
curation_stats <- data.frame(Method=method,Level=level,
                                Curated=processed,
                             perfect_match_share_mean, perfect_match_share_sd, 
                             refinds_share_mean, refinds_share_sd,
                             unknown_share_mean,unknown_share_sd,
                             redundant_share_mean, redundant_share_sd,
                             imperfect_share_mean, imperfect_share_sd,
                             refinds_lost_mean,refinds_lost_sd)

tab_name <- file.path(main_path,"Table_plot_curation_statistics_revision3.txt")
{write.table(curation_stats, tab_name, sep="\t",quote=FALSE, col.names = NA)}


tab_name <- file.path(main_path,"Table_plot_curation_statistics_revision3.txt")
plot_curation_statistics <- read.table(tab_name, sep="\t", header=TRUE, as.is=TRUE)

plot_curation_statistics$Level <- 
 factor(plot_curation_statistics$Level,levels = c("99/100", "98.5", "98", "97", 
                                           "96", "95"))
plot_curation_statistics$Method <- 
 factor(plot_curation_statistics$Method,levels = c("CROP","DADA2(+VS)","SWARM","VSEARCH"))
plot_curation_statistics$Curated <- 
 factor(plot_curation_statistics$Curated,levels = c("raw","curated"))

plot_curation_statistics <- plot_curation_statistics[with(plot_curation_statistics, order(Curated,Method,Level)),]

plot_curation_statistics[is.na(plot_curation_statistics)] <- 0

t_method <- plot_curation_statistics$Method[1:20]
t_level <- paste0(plot_curation_statistics$Level[1:20],"%")

t_refinds <- paste0(as.character(round(plot_curation_statistics$refinds_share_mean[1:20],2)), "±", 
                    as.character(round(plot_curation_statistics$refinds_share_sd[1:20],2)) , " / ",
                as.character(round(plot_curation_statistics$refinds_share_mean[21:40],2)), "±",
                as.character(round(plot_curation_statistics$refinds_share_sd[21:40],2)))

t_unknown <- paste0(as.character(round(plot_curation_statistics$unknown_share_mean[1:20],2)), "±", 
                    as.character(round(plot_curation_statistics$unknown_share_sd[1:20],2)) , " / ",
                as.character(round(plot_curation_statistics$unknown_share_mean[21:40],2)), "±",
                as.character(round(plot_curation_statistics$unknown_share_sd[21:40],2)))

t_redundant <- paste0(as.character(round(plot_curation_statistics$redundant_share_mean[1:20],2)), "±", 
                    as.character(round(plot_curation_statistics$redundant_share_sd[1:20],2)) , " / ",
                as.character(round(plot_curation_statistics$redundant_share_mean[21:40],2)), "±",
                as.character(round(plot_curation_statistics$redundant_share_sd[21:40],2)))

t_imperfect <- paste0(as.character(round(plot_curation_statistics$imperfect_share_mean[1:20],2)), "±", 
                    as.character(round(plot_curation_statistics$imperfect_share_sd[1:20],2)) , " / ",
                as.character(round(plot_curation_statistics$imperfect_share_mean[21:40],2)), "±",
                as.character(round(plot_curation_statistics$imperfect_share_sd[21:40],2)))

t_refinds_lost <- paste0(as.character(round(plot_curation_statistics$refinds_lost_mean[1:20],2)), "±", 
                    as.character(round(plot_curation_statistics$refinds_lost_sd[1:20],2)))

main_table <- data.frame(Method=t_method, Level=t_level, Imperfect_matches=t_imperfect, Refinds=t_refinds, Unknown=t_unknown, Redundant=t_redundant, Lost=t_refinds_lost)

tab_name <- file.path(main_path,"Table2_rev2x.txt")
{write.table(main_table, tab_name, sep="\t",quote=FALSE, col.names = NA)}

```

```{r}
#Not used. Can be used to evaluate the specific species being lost and kept pr site.
lost_species <- list()
kept_species <- list()
for (o in 1:20){
 for (i in 1:130){
 lost_species[[i]] <- refound_species[[o]][[i]][!refound_species[[o]][[i]] %in% refound_species[[o]][[i]]]
 kept_species[[i]] <- refound_species[[o]][[i]][refound_species[[o]][[i]] %in% refound_species[[o]][[i]]]
 }
}
```

